Portfolio prioritization of salmon conservation

### 2013-03-19 inst/examples/salmonport.tex
Updating files before major work after Jon, Andy, Trevor meetings

% my modifications:
\usepackage{fontspec}
\usepackage{booktabs}
\defaultfontfeatures{Ligatures=TeX}
\setsansfont{Calibri}
\setmonofont{Inconsolata}
\setmainfont{Linux Libertine}
\DeclareTextCommandDefault{\nobreakspace}{\leavevmode\nobreak\ }
\widowpenalty=1000
\clubpenalty=1000
\usepackage{setspace}
\onehalfspacing

% A bit bigger than one inch margins:
%\topmargin -1.43cm
%\oddsidemargin 0.4cm
%\evensidemargin 0.4cm
%\textwidth 14.5cm
\textheight 22.5cm

%\usepackage[left=1.5in,top=1.5in,right=1.5in,bottom=1.5in,nohead,nofoot]{geometry}
\usepackage{geometry}
\geometry{letterpaper}

@@

%\usepackage[T1]{fontenc}
%\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
%\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
%\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  %\usepackage[utf8]{inputenc}
%%\else % if luatex or xelatex
  %\usepackage{fontspec}
  %\ifxetex
    %\usepackage{xltxtra,xunicode}
  %\fi
  %\defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  %\newcommand{\euro}{€}
%%%%%\fi
% use microtype if available
%\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
%%%\usepackage{natbib}
%\bibliographystyle{plainnat}
%%%%%\usepackage{graphicx}
% We will generate all images so they have a width \maxwidth. This means
% that they will get their normal width if they fit onto the page, but
% are scaled down if they would overflow the margins.
%\makeatletter
%\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth
%\else\Gin@nat@width\fi}
%\makeatother
%\let\Oldincludegraphics\includegraphics
%\renewcommand{\includegraphics}[1]{\Oldincludegraphics[width=\maxwidth]{#1}}
%\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
%
\title{Portfolio prioritization of salmon conservation}
\author{Sean Anderson}
\date{}

@@

{[}Based partly on Jon's notes.{]} Managing risk is fundamental to the
conservation of an endangered species. When an endangered species exists
as a metapopulation, we can manage risk at two levels: at the population
level or at the metapopulation level. Typically we treat sources of risk
at the metapopulation level as external and uncontrollable and so we
manage risk by altering anthropogenic disturbances such as fishing or
hunting on a population level as well as improving connectivity of
populations.

The management of financial portfolios provides another way of
considering risk. Economists consider the risk and performance of a
financial portfolio as depending on the weighting of individual assets
that make up the portfolio. Modern Portfolio Theory (MPT) proposes that
there is a set of portfolios that achieves maximum expected return for a
level of expected risk or minimum expected risk for a level of expected
return \citep{Markowitz1952, Markowitz1959}. This optimal set contains
portfolios that range along a continuum of risk-tolerance; economists
refer to this set as the efficient frontier.

Similarly, expected growth rate and variance at a metapopulation level
will be a function of the properties of the populations. A portfolio
approach to managing risk for a metapopulation therefore considers how
conservation might affect the
\texttt{weight'' of each population in the metapopulation}portfolio''.

@@

Here, we conduct a post-hoc assessment of the set of optimal salmon
portfolios given various return-risk definitions and objectives. We ask:
What does a portfolio approach to management tell us about salmonid
conservation priorities? Does it produce any rules of thumb about what
optimal portfolio management of salmonid metapopulations would entail?

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  What are the attributes of the populations that form the optimal set?
  For example, are populations with high response diversity prioritized?
\item
  How much lower is stream-level and ESU-level extinction risk for the
  optimal portfolios than the less optimal portfolios? Extinction
  probability could be estimated through long-term simulation or through
  an asymmetric risk metric like conditional value at risk (CVaR). Given
  I am including straying, this will most likely be a
  quasi-extinction-risk metric.
\item
  How do our answers to other questions change depending on what we
  measure on the risk-return axes? For example, we could trade off the
  mean rate of change of salmon returns vs.~variance of the rate of
  change. This would be most similar to a financial portfolio
  optimization. We could also trade off the salmon returns (or catches)
  themselves. Further, we could work with variance or a downside risk
  metric like CVaR.
\item
  How do constraints (e.g.~viable salmonid population (VSP)-like
  constraints) affect the selected portfolios in risk-return space and
  in terms of common metrics such as population-level and ESU-level
  extinction risk?
\item
  How do the properties of the optimal portfolio set change based on
  different kinds of environmental driver patterns? For example,
  magnitude of driver variance, cyclical dynamics like the PDO, black
  swans or unexpected surprises, increases in temperature and
  variability through time (climate change).
\end{enumerate}

@@

\begin{figure}[htbp]
  \centering
    \includegraphics[height=7in]{simulation_diagram.pdf}
 \caption{Flow chart of the salmon-metapopulation simulation and portfolio optimization. There are $n$ salmon populations and $t$ generations. Blue text indicates values that are generated before the simulation progresses through time. Orange text indicates steps in which calculations are performed. Black text indicates values that are calculated. Grey text indicates parameters that can be set. Green text indicates the looping structure of the simulation.}
  \label{fig:sim-diagram}
\end{figure}

@@

\noindent
where $t$ represents a generation time, $i$ represents a population, $R$
is the number of returns, $S$ is the number of spawners, $a$ is the
productivity parameter (which can vary with the environmental signal),
and $b$ is the density-dependent term (which is used as the asset
weights in the portfolios). The term $w_{ti}$ represents first-order
autocorrelated error (AR1). Formally, $w_{ti} = w_{ti-1} \rho + v_{ti}$,
where $v_{ti}$ represents independent and normally distributed error
with mean 0 and standard deviation of $\sigma_v$. The parameter $\rho$
represents the correlation between residuals from subsequent years.

@@

I have programmed five different kinds of environmental signals. I think
it's easiest to think of these as temperature signals, but they could
represent more general environmental anomalies, such as the PDO. The
five kinds of environmental signals are described in Table
\ref{tab:env-types} and illustrated in Figure \ref{fig:env-ts}. Later, I
may include black-swan-like events, although the regime shifts could be
seen this way if the changes are dramatic enough. So far all figures use
a sine wave. The problem with using a randomly generated ARMA time
series is that you get much greater differences in the portfolios from
run to run.

\begin{table}[h!]
  \centering
  \small
  \caption{Environmental signals, parameters, and descriptions.}
\begin{tabular}{p{3.2cm}p{3.25cm}p{6.8cm}}
\toprule
Signal type & Parameters & Description \\
\midrule
Sine wave & $\overline{T} + A \cdot \sin(\omega t + \phi)$& $\overline{T}$ is 
the mean environmental value, $A$ is the amplitude, $\omega$ is the frequency in 
radians, and $\phi$ is the phase\\
ARMA & AR1, MA1, $\sigma_T$ & First-order autoregressive (AR1) and moving 
average (MA1) parameters and standard deviation; mean is assumed to be 0\\
Regime shifts & $T_{b0},T_{b1}, \ldots, T_{bn}$; $T_{v0},T_{v1}, \ldots, T_{vn}$ 
& Breakpoints ($T_{b}$) and values ($T_v$) for regimes $1$ through $n$\\
Linear change & $T_{\mathrm{min}}$, $T_{\mathrm{max}}$ & Minimum and maximum 
values\\
Constant & $T_{\mathrm{constant}}$& Constant value\\
\bottomrule
\end{tabular}
\label{tab:env-types}
\end{table}

@@

I've implemented straying as in \citet{Cooper1999}. I generate a matrix
that represents the fraction of straying between any two populations. I
have set up the metapopulation (thus far) as in in a very simple
scenario: the populations are arranged in a line and those that are
nearer to each other are more likely to stray between each other
{[}insert caveats{]}. Two parameters control the straying: the fraction
of fish that stray from their natal stream in any given generation $f$
and the rate at which this straying between streams decays with distance
$m$. I calculated the number of salmon straying from stream $i$ to
stream $j$ as,

\begin{equation}
  \mathrm{strays}_{tij} = \frac{S_{ti} f e^{-m \lvert i-j \rvert }}{\sum\limits_{k \neq j}^{} e^{-m \lvert k-j \rvert }},
\end{equation}

where the denominator is a normalization constant to ensure that no
salmon are gained or lost. The subscript $k$ represents a stream ID.
{[}Not sure this is written correctly --- I'll think about it --- but it
should be coded correctly.{]} There is no straying from and to the same
stream. See Figure \ref{fig:stray-matrix} for an example straying
matrix.

@@

where $a_i^*$ is the maximum possible productivity parameter for
population $i$, $T_t$ is the temperature at time step $t$, $T_i^*$ is
the optimal temperature for population $i$, and $\gamma_i$ controls the
width of the thermal tolerance curve for each population. See Figure
\ref{fig:thermal-curves} for the default thermal tolerance curves for
ten populations.

@@

\begin{figure}[htbp]
\centering
\includegraphics{figure/thermal-curves.pdf}
\caption{Thermal tolerance curves for ten populations. Here, and
throughout this document, I have used the same rainbow of colours to
identify the salmon populations. The populations are arranged from blue
(more productive with cooler temperatures) to red (more productive with
warmer temperatures\label{fig:thermal-curves}}
\end{figure}

@@

\begin{verbatim}
  n_t = 120, # number of years
  n_pop = 10, # number of populations
  stray_decay_rate = 0.3, # rate that straying decays with distance
  stray_fraction = 0.01, # fraction of fish that stray from natal streams
  b = rep(1000, n_pop), # Ricker density-dependent parameter
  spawners_0 = round(b), # spawners at start
  sigma_v = 0.1, # stock-recruit residual SD
  v_rho = 0.4, # stock-recruit residual AR1 correlation
  a_width_param = rep(0.02, n_pop), # width of thermal curves by pop
  optim_temp = seq(13, 19, length.out = n_pop), # optimal temperatures by pop
  max_a = rep(1.4, n_pop), # maximum Ricker a values by pop at optimum temp
  env_type = c("sine", "arma", "regime", "linear", "constant"),
  env_params = list(amplitude = 2.0, ang_frequency = 0.2, phase = 0, mean_value = 16),
  use_cache = FALSE, # regenerate stochastic values? 
  add_straying = TRUE # include or ignore straying
\end{verbatim}

@@

\begin{verbatim}
Error: could not find function "ricker_ar1"
\end{verbatim}

@@

\begin{verbatim}
Error: could not find function "ricker_ar1"
\end{verbatim}

@@

>>>>

% my modifications:
\usepackage{fontspec}
\usepackage{booktabs}
\usepackage{mathptmx}
\defaultfontfeatures{Ligatures=TeX}
\setsansfont{Calibri}
\setmonofont{Inconsolata}
%\setmainfont{Linux Libertine}
%\setmainfont{Times New Roman}
\setmainfont{Calibri}
%\setmainfont{Helvetica}
\DeclareTextCommandDefault{\nobreakspace}{\leavevmode\nobreak\ }
\widowpenalty=1000
\clubpenalty=1000
\usepackage{setspace}
\onehalfspacing
%\doublespacing

\usepackage{titlesec}
\titleformat*{\section}{\Large\bfseries\sffamily}
\titleformat*{\subsection}{\large\sc\sffamily}
\titleformat*{\subsubsection}{\itshape\sffamily}

\usepackage{geometry}
\geometry{letterpaper}

@@

%\usepackage[T1]{fontenc}
%\usepackage{lmodern}

% end customization except for commenting out lines
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
%\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
%\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  %\usepackage[utf8]{inputenc}
%%\else % if luatex or xelatex
  %\usepackage{fontspec}
  %\ifxetex
    %\usepackage{xltxtra,xunicode}
  %\fi
  %\defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  %\newcommand{\euro}{€}
%%%%%\fi
% use microtype if available
%\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
%%%\usepackage{natbib}
%\bibliographystyle{plainnat}
%%%%%\usepackage{graphicx}
% We will generate all images so they have a width \maxwidth. This means
% that they will get their normal width if they fit onto the page, but
% are scaled down if they would overflow the margins.
%\makeatletter
%\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth
%\else\Gin@nat@width\fi}
%\makeatother
%\let\Oldincludegraphics\includegraphics
%\renewcommand{\includegraphics}[1]{\Oldincludegraphics[width=\maxwidth]{#1}}
%\urlstyle{same}  % don't use monospace font for urls
%\setlength{\parindent}{0pt}
%\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
%
\title{Portfolio prioritization of salmon conservation}
\author{Sean Anderson}
\date{}

@@

Managing risk is fundamental to the conservation of an endangered
species. When an endangered species exists as a metapopulation, we can
manage risk at two levels: at the population level or at the
metapopulation level. Typically we treat sources of risk at the
metapopulation level as external and uncontrollable and so we manage
risk by altering fishing or hunting on a population level as well as
improving the connectivity of populations.

The management of financial portfolios provides another way of
considering risk. Economists consider the risk and performance of a
financial portfolio as a function of the weighting of individual assets
that make up the portfolio. Modern Portfolio Theory (MPT) proposes that
there is a set of portfolios that maximizes expected return for a level
of expected risk or minimizes expected risk for a level of expected
return \citep{Markowitz1952, Markowitz1959}. This optimal set contains
portfolios that range along a continuum of risk-tolerance; economists
refer to this set as the efficient frontier.

Similarly, expected growth rate and variance at a metapopulation level
is a function of the variance, covariance, and size of the individual
populations. A portfolio approach to managing risk for a metapopulation
therefore considers how conservation might affect the ``weight'' of each
population in a metapopulation ``portfolio''.

@@

The basic goal is to conduct a post-hoc assessment of the set of optimal
salmon portfolios given various return-risk definitions and objectives.
We ask: What does a portfolio approach to management tell us about
salmonid conservation priorities? Can we produce any rules of thumb
about what optimal portfolio management of salmonid metapopulations
would entail?

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  What are the attributes of the populations that form the optimal set?
  For example, does the optimal set tend to focus on populations with as
  diverse responses to the environment as possible?
\item
  How much lower is stream-level and ESU-level extinction risk for the
  optimal portfolios than the less optimal portfolios? Extinction
  probability could be estimated through long-term simulation or through
  an asymmetric risk metric like conditional value at risk (CVaR). Given
  I am including straying, this will most likely be a
  quasi-extinction-risk metric.
\item
  How do our answers to other questions change depending on what we
  measure on the risk-return axes? For example, we could trade off the
  mean rate of change of salmon returns vs.~variance of the rate of
  change. This would be most similar to a financial portfolio
  optimization. We could also trade off the salmon returns (or catches)
  themselves. Further, we could work with variance or a downside risk
  metric like CVaR.
\item
  How do constraints (e.g.~viable salmonid population (VSP)-like
  constraints) affect the selected portfolios in risk-return space and
  in terms of common metrics such as population-level and ESU-level
  extinction risk?
\item
  How do the properties of the optimal portfolio set change based on
  different kinds of environmental driver patterns? For example,
  magnitude of driver variance, cyclical dynamics like the PDO, black
  swans or unexpected surprises, increases in temperature and
  variability through time (climate change).
\end{enumerate}

@@

\begin{figure}[htbp]
  \centering
    \includegraphics[height=6.5in]{simulation_diagram.pdf}
 \caption{Flow chart of the salmon-metapopulation simulation and portfolio optimization. There are $n$ salmon populations and $t$ generations. Blue text indicates values that are generated before the simulation progresses through time. Orange text indicates steps in which calculations are performed. Black text indicates values that are calculated. Grey text indicates parameters that can be set. Green text indicates the looping structure of the simulation.}
  \label{fig:sim-diagram}
\end{figure}

@@

\noindent
where $t$ represents a generation time, $i$ represents a population, $R$
is the number of returns, $S$ is the number of spawners, $a$ is the
productivity parameter (which can vary with the environmental signal),
and $b$ is the density-dependent term (which is used as the asset
weights in the portfolios). The term $w_{ti}$ represents first-order
autocorrelated error (AR1). Formally, $w_{ti} = w_{ti-1} \rho + v_{ti}$,
where $v_{ti}$ represents independent and normally distributed error
with mean 0 and standard deviation of $\sigma_v$. The parameter $\rho$
represents the correlation between residuals from subsequent
generations.

@@

I have programmed five different kinds of environmental signals. I think
it's easiest to think of these as temperature, but they could represent
more general environmental anomalies, such as the PDO. In Table
\ref{tab:env-types}, I describe the five kinds of environmental signals
I have included. Examples of these time series are illustrated in Figure
\ref{fig:env-ts}. Later, I may include black-swan-like events, although
the regime shifts could be seen this way if the changes are dramatic
enough. So far, all simulation results in the rest of this document use
a sine wave. The problem with using a randomly generated ARMA time
series is that you get much greater differences in the results from run
to run. This is fine in the long run, but hard to detect patterns
initially.

\begin{table}[h!]
\centering
\small
\caption{Environmental signals, parameters, and descriptions.}
\begin{tabular}{p{3.2cm}p{3.25cm}p{6.8cm}}
\toprule
Signal type & Parameters & Description \\
\midrule
Sine wave & $\overline{T} + A \cdot \sin(\omega t + \phi)$& $\overline{T}$ is 
the mean environmental value, $A$ is the amplitude, $\omega$ is the frequency in 
radians, and $\phi$ is the phase\\
ARMA & AR1, MA1, $\sigma_T$ & First-order autoregressive (AR1) and moving 
average (MA1) parameters and standard deviation; mean is assumed to be 0\\
Regime shifts & $T_{b0},T_{b1}, \ldots, T_{br-1}$; $T_{v0},T_{v1}, \ldots, T_{vr}$ 
& Breakpoints ($T_{b}$) and values ($T_v$) for regimes $1$ through $r$\\
Linear change & $T_{\mathrm{min}}$, $T_{\mathrm{max}}$ & Minimum and maximum 
values\\
Constant & $T_{\mathrm{constant}}$& Constant value\\
\bottomrule
\end{tabular}
\label{tab:env-types}
\end{table}

@@

I've implemented straying as in \citet{Cooper1999}. I generate a matrix
that represents the fraction of straying between any two populations. I
have set up the metapopulation (thus far) in a very simple scenario: the
populations are arranged in a line and those that are nearer to each
other are more likely to stray between each other {[}insert caveats{]}.
Two parameters control the straying: the fraction of fish $f$ that stray
from their natal stream in any given generation and the rate at which
this straying between streams decays with distance $m$. I calculated the
number of salmon straying from stream $j$ to stream $i$ as,

\begin{equation}
  \mathrm{strays}_{tij} = f R_{tj} \frac{e^{-m \lvert i-j \rvert }}{\displaystyle\sum\limits_{\substack{k = 1 \\
  k \neq j}}^{n} e^{-m \lvert k-j \rvert }},
\end{equation}

\noindent
where $R_{tj}$ is the number of returning salmon at generation $t$ whose
natal stream was stream $j$. The subscript $k$ represents a stream ID
and $n$ the number of populations. The denominator is a normalizing
constant to ensure the desired fraction of fish stray. See Figure
\ref{fig:stray-matrix} for an example straying matrix.

@@

\noindent
where $a_i^*$ is the maximum possible productivity parameter for
population $i$, $T_t$ is the temperature at time step $t$, $T_i^*$ is
the optimal temperature for population $i$, and $\gamma_i$ controls the
width of the thermal tolerance curve for each population. See Figure
\ref{fig:thermal-curves} for the default thermal tolerance curves for
ten populations.

@@

\begin{figure}[htbp]
\centering
\includegraphics{figure/thermal-curves.pdf}
\caption{Thermal tolerance curves for ten populations. Here, and
throughout this document, I have used the same rainbow of colours to
identify the salmon populations. The populations are arranged from blue
(more productive with cooler temperatures) to red (more productive with
warmer temperatures).\label{fig:thermal-curves}}
\end{figure}

@@

\begin{verbatim}
n_t = 120, # number of generations
n_pop = 10, # number of populations
stray_decay_rate = 0.3, # rate that straying decays with distance
stray_fraction = 0.01, # fraction of fish that stray from natal streams
b = rep(1000, n_pop), # Ricker density-dependent parameter
spawners_0 = round(b), # spawners at start
sigma_v = 0.1, # stock-recruit residual SD
v_rho = 0.4, # stock-recruit residual AR1 correlation
a_width_param = rep(0.02, n_pop), # width of thermal curves by pop
optim_temp = seq(13, 19, length.out = n_pop), # optimal temperatures by pop
max_a = rep(1.4, n_pop), # maximum Ricker a values by pop at optimum temp
env_type = c("sine", "arma", "regime", "linear", "constant"),
env_params = list(amplitude = 2.0, ang_frequency = 0.2, phase = 0, mean_value = 16),
use_cache = FALSE, # regenerate stochastic values? 
add_straying = TRUE # include or ignore straying
\end{verbatim}

@@

\begin{figure}[htbp]
\centering
\includegraphics{figure/plot-base-rickers.pdf}
\caption{We can also look at samples of the Ricker curves by population
(population numbers are in the upper left corners). Here, I've sampled
40 of the $a$ parameters through time for each population. The shaded
rectangles to indicate the range of spawners observed in the simulation.
The colour of the line indicates the temperature anomaly in that year
with blue indicating cooler years and red indicating warmer years. In
this example all the populations have the same $b$ parameters. Notice
how the productivity is more variable for populations nearer the limits
of their thermal tolerance (e.g.~population 1 vs.~population 5). Also
notice how population 1 is more productive in cool years whereas
population 10 is more productive in warm years.}
\end{figure}

@@

\begin{figure}[htbp]
\centering
\includegraphics{figure/unnamed-chunk-1.pdf}
\caption{Example Ricker curves for the minimum-variance metapopulation.}
\end{figure}

@@

\begin{figure}[htbp]
\centering
\includegraphics{figure/plot-weights-samples.pdf}
\caption{A sample of 30 minimum-variance portfolios given the same set
of starting parameters as for the example illustrated above. The dashed
grey line indicates the weights if all populations had an equal carrying
capacity. As it stands, the weights look fairly random to me.}
\end{figure}

\begin{figure}[htbp]
\centering
\includegraphics{figure/plot-weights-median.pdf}
\caption{Median (solid line) and 25th and 75th quantiles of the weights
across 40 minimum-variance portfolios. Just another way of looking at
it.}
\end{figure}
